---
title: "Datavyur Package Tutorial"
author: "Joseph M. Burling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Datavyur Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(eval=FALSE)
library(datavyur)
```


# Basic `datavyur` usage

This section will briefly outline how to use some of the functions found in the `datavyur` package.
To follow along, make sure you know where your `.csv` files have been saved when you used the `datavyu2csv.rb` script in the previous step. If you don't have `.csv` files to work with, you can use the default ones built into this package.

First, you need to load the `datavyur` library. Make sure it has already been installed before loading. If you're reading this on GitHub, go here to install: 

[How to install datavyur](https://github.com/iamamutt/datavyu#how-to-install)

Use the R function `library` to load the `datvyur` package after it has been installed

```{r}
# Load the datavyur library to use the functions below
library(datavyur)
```

You have to then let R know where your `datavyur` data is at. This is the specially formatted `.csv` files obtained when using the Ruby script `datavyu2csv.rb`, as seen in this tutorial:

(Getting started with Datavyu)[https://github.com/iamamutt/datavyu/blob/master/tutorial/datavyu_tutorial.md]

To set the default path for the data that was exported, override the default option `datavyur.folder` by doing the following in the RStudio console:

```{r}
# Replace the string path/to/data with the location to where your data is located, relative to working directory.
options(datavyur.folder="path/to/data")
```

By default, the path is set to use the data built into this package. The code below will tell you where `datavyur` is searching for Datavyu files.

```{r}
getOption("datavyur.folder")
```

An alternative to setting an R option is to just create an R object that holds the path string, like so:

```{r}
# Find the full path to where the .csv files have been saved
# Replace path/to/data with the correct path
data_path <- normalizePath("path/to/data")
```

You can then use this path as an argument for some of the functions, like when using `import_column` or `datavyu_col_search`, as you'll see later.

The `datavyur` package provides several utility functions to simplify the data manipulation process. In the command window you can use the `?` operator to see the manual for each function. For example, to see the manual for the `import_column` function type the following in the command window:

```{r}
# see help documentation
?import_column
```

Type in the search below to see a list of functions available in the `datavyur` package.

```{r}
# see help documentation
?datavyur
```

## Viewing column information

The function `datavyu_col_search` will search through the path you specify and find all Datavyu files that have been already exported to `.csv` files, along with their column names and other attributes.

```{r}
# if datavyur.folder option has been set
datavyu_col_search()

# otherwise use explicit folder path stored in data_path
datavyu_col_search(data_path)
```

```{r, echo=FALSE, eval=TRUE, results='asis'}
# if datavyur.folder option has been set
fdat <- datavyu_col_search()
fdat$local <- "./*.csv"
knitr::kable(fdat)
```

To view only the column names from each file:

```{r}
# if datavyur.folder option has been set
datavyu_col_search(unq=TRUE, cnames="column")
```

```{r, echo=FALSE, eval=TRUE, results='asis'}
knitr::kable(datavyu_col_search(unq=TRUE, cnames="column"))
```


## Importing Datavyu columns into R

The ruby script `datavyu2csv.rb` exports a `.csv` file for each column within each file. To combine them back together in R you'll use the function `import_column`. You'll need to know the path to your data again if not already set, and the names of the columns you're trying to import (using the `datavyu_col_search`, if you don't already know this).

The code below will search the example `.csv` and look for columns with a specific name, them load them into R. This is assuming the folder path has already been set above in the options. If not, set the `folder` argument from this function to a path pointing to the data folder.

```{r, eval=TRUE}
# load columns as separate data frames
child_hands <- import_column("childhands") # or import_column("childhands", data_path)
parent_hands <- import_column("parenthands")
```

You can also append the datavyu column name to the data using the argument `append.colnames` and setting it to `TRUE`.

```{r, eval=TRUE}
parent_hands <- import_column("parenthands", append.colnames = TRUE)
```


partial data obtained from `child_hands`:

```{r, echo=FALSE, eval=TRUE, results='asis'}
knitr::kable(head(child_hands))
```

partial data obtained from `parent_hands` with `append.colnames = TRUE`:

```{r, echo=FALSE, eval=TRUE, results='asis'}
knitr::kable(head(parent_hands))
```

## Merging Nested Columns

If you have columns that are originally nested within each other, you can get it back to this format by using the function `merge_nested`. The example below will attempt to merge two data frames by timestamps (even though these data aren't really nested).

Since these data aren't really nested, you'll get a lot of rows that won't merge. But you can still include them if you want. This is the default behavior of the function.

```{r}
z1 <- merge_nested(child_hands, parent_hands)
```

The data `z1` above merges two data frames by column and includes any rows that couldn't be merged. If you see `NA` for any rows corresponding to `column.1` (the higher level), this means that timestamps were found for the lower level data but not for the higher level. This should not happen if you're data are truly nested. Fix any errors in your original `.opf` file if this is the case. An `NA` for `column.2` means that higher level timestamps exist, but no lower level data. This is okay since sometimes no observations exists for some higher level row.

To merge without all the `NA` values, do the following. I'm also specifying different suffixes besides `.1` and `.2`, and setting the argument `keepall=FALSE` to remove all `NA`s.

```{r}
z2 <- merge_nested(child_hands, parent_hands, ids=c(".higher", ".lower"), keepall=FALSE)
```

## Checking for invalid codes and timestamps


## R Data to Spreadsheet

Save any R data frame to a spreadsheet that can be opened in Excel. You don't need the `datavyur` packages for this, and comes with standard R. You can do this for any R data frame.

```{r}
write.csv(z2, file="merged_data.csv", row.names=FALSE, na="")

```

# Advanced usage

## Align data by time or by cell number (ordinal)



## R Data to Datavyu

If you have R data that you want to convert back into something that the Datavyu program can recognize, you can do so with the function `r2datavyu`. You just need to make sure you have the columns `ordinal`, `onset`, and `offset` in your R data. You can pass all data frames you want to convert as a single `list` object. Each object in the list must be named (list items must have names)! Or the conversion to `.opf` will not work.

```{r}
# provide a list of data to convert
r2datavyu(list(chands=child_hands, phands=parent_hands), "myexport")
```

The function above will save the file as a weirdly formatted `.csv` file, but this is something that can be used in Datavyu, but not right away. Since the Datavyu program doesn't allow you to directly import these types of `.csv` files using the GUI, you need to use a script. Luckily I have provided one for you in the `general` folder from the github repository.

In Datavyu, run the script called `csv2opf.rb`. This will convert all `.csv` files (properly formmated by using the function `r2datavyu`) in a folder to `.opf` files, which can now be opened in Datavyu directly.


# Other useful functions

## Making fake data

The `datavyur` package provides a function to create fake data called `datavyu_dat`. Below I'm just creating two separate datasets using this function. This is handy for seeing how your data should be formatted if you want to save an R data back into Datavyu data.

```{r}
x <- as.data.frame(datavyu_dat(n1=25, n2=2)[[1]])
y <- datavyu_dat(n1=2, n2=100)[[2]]
```

## Time Conversion

Datavyu prints timestamps in milliseconds. You can convert this to a more readable format by using the function `ms2time`.

```{r}
# print milliseconds to time string
ms2time(x$onset)
```

You can also save this conversion back into your R data if you like.

```{r}
# save time string back into data frame
x$onset_str <- ms2time(x$onset)
x$offset_str <- ms2time(x$offset)
```

## Timestamps to frame numbers


